---
title: "betydata"
subtitle: "Public Data from the BETYdb Database"
---

```{r}
#| echo: false
library(betydata)
source("R/homepage-data.R", local = TRUE)
```

**betydata** is an R data package providing offline access to public data from [BETYdb](https://betydb.org) (Biofuel Ecophysiological Traits and Yields database). It enables reproducible analyses of plant traits, crop yields, and supporting metadata without requiring database connectivity.

## Guides

| Guide | Description |
|-------|-------------|
| [Getting Started](vignettes/getting_started.qmd) | Package overview, data model, and key concepts |
| [Common Analyses](vignettes/common_analyses.qmd) | Practical analysis patterns with dplyr |
| [PFTs & Priors](vignettes/pfts-priors.qmd) | Plant Functional Types and Bayesian prior distributions |
| [Manuscript Reproduction](vignettes/manuscript.qmd) | Reproduce key analyses from LeBauer et al. (2018) |

## At a Glance

```{r}
#| echo: false
#| results: asis
metrics <- homepage_metrics(traitsview)

cat("::: {.homepage-metrics}\n")
for (i in seq_len(nrow(metrics))) {
  cat("::: {.homepage-metric-card}\n")
  cat(sprintf("### %s\n\n", metrics$metric[[i]]))
  cat(sprintf("<span class=\"homepage-metric-value\">%s</span>\n\n", format(metrics$value[[i]], big.mark = ",")))
  cat(sprintf("%s\n", metrics$definition[[i]]))
  cat(":::\n")
}
cat(":::\n")
```

## Installation

```r
# install.packages("remotes")
remotes::install_github("PecanProject/betydata")
```

## Quick Start

```{r}
#| echo: false
if (!requireNamespace("DT", quietly = TRUE)) {
  stop("Package 'DT' is required to render the searchable homepage preview.", call. = FALSE)
}

preview <- traitsview[, c(
  "trait", "mean", "units", "scientificname", "sitename",
  "author", "citation_year", "year"
)]
names(preview) <- c(
  "Trait", "Value", "Units", "Species", "Site",
  "Data-source author", "Citation year", "Observation year"
)

DT::datatable(
  preview,
  caption = "Searchable preview of trait and yield observations",
  rownames = FALSE,
  options = list(
    deferRender = TRUE,
    pageLength = 25,
    lengthMenu = c(10, 25, 50, 100),
    scrollX = TRUE
  )
)
```

## Most Represented Variables and Species

```{r}
#| echo: false
#| label: tbl-top-traits
#| tbl-cap: "Variables with the most observation records"
homepage_top_traits(traitsview) |>
  knitr::kable(
    col.names = c("Variable", "Description", "Observations"),
    row.names = FALSE
  )
```

```{r}
#| echo: false
#| label: tbl-top-species
#| tbl-cap: "Species with the most observation records"
homepage_top_species(traitsview)[
  , c("scientific_name", "common_name", "observations")
] |>
  knitr::kable(
    col.names = c("Scientific name", "Common name", "Observations"),
    row.names = FALSE
  )
```

## Data License

Data: [ODC-By-1.0](https://opendatacommons.org/licenses/by/1-0/)
Code: [BSD-3-Clause](https://opensource.org/licenses/BSD-3-Clause)

## Citation

LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)
