---
title: "Reproducing BETYdb Manuscript Analyses"
vignette: >
  %\VignetteIndexEntry{Reproducing BETYdb Manuscript Analyses}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = requireNamespace("betydata", quietly = TRUE))
```

::: {.callout-note}
## What you will learn

- How to reproduce key figures and tables from the BETYdb publication
- How to summarize trait and yield data by genus, trait, and site
- How current data compares to the 2017 snapshot used in the paper
:::

## Introduction

This vignette reproduces key analyses from the BETYdb manuscript (LeBauer et al., 2018) using the `betydata` package.

**Citation:** LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)

## Setup

```{r}
library(betydata)
library(dplyr)
library(ggplot2)

theme_set(theme_bw(base_size = 10, base_family = "sans"))
```

## Figure 1: Data Summary by Genus {#sec-fig1}

The manuscript presents trait and yield counts for key bioenergy genera. We reproduce this using `traitsview`:

```{r}
#| label: tbl-genus-summary
#| tbl-cap: "Trait and yield counts by bioenergy genus"
bioenergy_genera <- c("Miscanthus", "Panicum", "Populus", "Saccharum",
                       "Pinus", "Salix", "Robinia")

genus_summary <- traitsview |>
  filter(genus %in% bioenergy_genera, checked >= 0) |>
  summarise(
    n_traits = sum(result_type == "traits", na.rm = TRUE),
    n_yields = sum(result_type == "yields", na.rm = TRUE),
    total = n(),
    .by = genus
  ) |>
  arrange(desc(total))

genus_summary |>
  knitr::kable(col.names = c("Genus", "Traits", "Yields", "Total"))
```

::: {.callout-important}
## Comparison Notes

Counts may differ from the published manuscript because:

1. **QA/QC filtering:** betydata excludes `checked = -1` (failed QA/QC records)
2. **Snapshot date:** betydata was exported from a current database snapshot; the manuscript used 2017 data
3. **Access level:** betydata includes only public data from BETYdb
:::

## Figure 2: Trait Records by Genus {#sec-fig2}

```{r}
#| label: fig-trait-by-genus
#| fig-cap: "Number of records for focal traits across bioenergy genera (log scale)"
#| fig-height: 6
focal_traits <- c("Ayield", "leafN", "LAI", "SLA", "Vcmax",
                   "leaf_respiration_rate_m2", "Jmax")

trait_counts <- traitsview |>
  filter(
    genus %in% bioenergy_genera,
    trait %in% focal_traits,
    checked >= 0
  ) |>
  count(genus, trait, name = "n")

ggplot(trait_counts, aes(x = genus, y = n, fill = trait)) +
  geom_col(position = "dodge") +
  scale_y_log10(breaks = c(1, 10, 100, 1000, 10000)) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Number of Records (log scale)",
    fill = "Trait"
  ) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

## Figure 3: Trait Distributions {#sec-fig3}

The manuscript displays histograms of trait values across genera, showing the spread of measured values for key ecophysiological parameters.

```{r}
#| label: fig-trait-distributions
#| fig-cap: "Distribution of trait values for bioenergy genera"
#| fig-height: 8
hist_traits <- c("Ayield", "SLA", "Vcmax", "LAI")

trait_data <- traitsview |>
  filter(
    trait %in% hist_traits,
    !is.na(mean),
    checked >= 0,
    genus %in% bioenergy_genera
  )

ggplot(trait_data, aes(x = mean, fill = genus)) +
  geom_histogram(bins = 25, alpha = 0.7) +
  facet_wrap(~trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed Value",
    y = "Count",
    fill = "Genus"
  ) +
  theme(
    legend.position = "bottom",
    strip.background = element_blank()
  )
```

## Table 1: Database Contents Summary {#sec-table1}

```{r}
#| label: tbl-contents
#| tbl-cap: "Data records by genus (top 15 with >= 100 records)"
contents <- traitsview |>
  filter(checked >= 0) |>
  summarise(
    n_traits = sum(result_type == "traits", na.rm = TRUE),
    n_yields = sum(result_type == "yields", na.rm = TRUE),
    total = n(),
    .by = genus
  ) |>
  filter(total >= 100) |>
  arrange(desc(total))

knitr::kable(
  head(contents, 15),
  col.names = c("Genus", "Traits", "Yields", "Total")
)
```

## Yield Meta-Analysis Subset {#sec-meta-analysis}

The manuscript includes a meta-analysis of *Miscanthus* and switchgrass (*Panicum*) yields. Here we extract the relevant subset and compute summary statistics.

```{r}
#| label: tbl-yield-ma
#| tbl-cap: "Yield meta-analysis summary for Miscanthus and Panicum"
yield_ma <- traitsview |>
  filter(
    genus %in% c("Miscanthus", "Panicum"),
    trait == "Ayield",
    !is.na(lat),
    !is.na(lon),
    !is.na(mean),
    checked >= 0
  ) |>
  select(
    id, genus, scientificname, mean, units,
    n, stat, statname, lat, lon,
    author, citation_year, sitename, site_id
  )

yield_ma |>
  summarise(
    n_records = n(),
    mean_yield = round(mean(mean), 1),
    sd_yield = round(sd(mean), 1),
    n_sites = n_distinct(site_id),
    .by = genus
  ) |>
  knitr::kable(col.names = c("Genus", "Records", "Mean Yield (Mg/ha)", "SD", "Sites"))
```

## Figure 4: Yield Responses to Management {#sec-fig4}

Figure 4 links the manuscript's focal grass yields to management records through
`managements_treatments`. The bridge table can associate several management
records with one treatment, so each exposure is first reduced to one
analysis-ready row per treatment before it is joined to a yield observation.

```{r}
#| label: data-fig4-management
grass_yields <- traitsview |>
  filter(
    result_type == "yields",
    trait == "Ayield",
    units == "Mg ha-1 yr-1",
    genus %in% c("Miscanthus", "Panicum"),
    checked >= 0,
    !is.na(mean),
    !is.na(treatment_id),
    !is.na(year)
  ) |>
  transmute(
    yield_id = id,
    treatment_id,
    genus,
    yield = mean,
    observation_year = year,
    site_id
  )

stopifnot(!anyDuplicated(grass_yields$yield_id))

management_details <- managements_treatments |>
  select(management_id, treatment_id) |>
  left_join(
    managements |>
      select(
        management_id = id,
        mgmttype,
        level,
        units,
        management_date = date
      ),
    by = "management_id",
    relationship = "many-to-one"
  )

nitrogen_by_treatment <- management_details |>
  filter(
    mgmttype == "fertilizer_N_rate",
    !is.na(treatment_id),
    !is.na(level),
    units %in% c("kg ha-1 y-1", "g m-2 y-1")
  ) |>
  mutate(
    nitrogen_rate = if_else(units == "g m-2 y-1", level * 10, level)
  ) |>
  summarise(
    nitrogen_events = n(),
    nitrogen_values = n_distinct(nitrogen_rate),
    nitrogen_rate = first(nitrogen_rate),
    .by = treatment_id
  ) |>
  filter(nitrogen_values == 1) |>
  select(treatment_id, nitrogen_rate, nitrogen_events)

planting_by_treatment <- management_details |>
  filter(mgmttype == "planting", !is.na(treatment_id)) |>
  mutate(planting_date = as.Date(management_date)) |>
  filter(!is.na(planting_date)) |>
  summarise(
    planting_date = min(planting_date),
    planting_records = n(),
    .by = treatment_id
  ) |>
  mutate(planting_year = as.integer(format(planting_date, "%Y"))) |>
  select(treatment_id, planting_year, planting_records)

stopifnot(
  !anyDuplicated(nitrogen_by_treatment$treatment_id),
  !anyDuplicated(planting_by_treatment$treatment_id)
)

yield_nitrogen <- grass_yields |>
  inner_join(
    nitrogen_by_treatment,
    by = "treatment_id",
    relationship = "many-to-one"
  ) |>
  filter(nitrogen_rate >= 0)

yield_age <- grass_yields |>
  inner_join(
    planting_by_treatment,
    by = "treatment_id",
    relationship = "many-to-one"
  ) |>
  mutate(stand_age = observation_year - planting_year) |>
  filter(stand_age >= 0)

stopifnot(
  !anyDuplicated(yield_nitrogen$yield_id),
  !anyDuplicated(yield_age$yield_id),
  all(yield_nitrogen$nitrogen_rate >= 0),
  all(yield_age$stand_age >= 0)
)
```

The nitrogen rate uses `kg ha-1 y-1` as the analysis unit; equivalent
`g m-2 y-1` records are converted before joining. Treatments with conflicting,
undated nitrogen rates are excluded rather than assigned an arbitrary rate.
Repeated records with the same rate are collapsed to one treatment row.

```{r}
#| label: tbl-fig4-coverage
#| tbl-cap: "Coverage of yield-management analyses"
analysis_coverage <- bind_rows(
  yield_nitrogen |>
    summarise(
      records = n(),
      treatments = n_distinct(treatment_id),
      sites = n_distinct(site_id),
      .by = genus
    ) |>
    mutate(analysis = "Nitrogen fertilization"),
  yield_age |>
    summarise(
      records = n(),
      treatments = n_distinct(treatment_id),
      sites = n_distinct(site_id),
      .by = genus
    ) |>
    mutate(analysis = "Stand age")
) |>
  select(analysis, genus, records, treatments, sites)

analysis_coverage |>
  knitr::kable(col.names = c("Analysis", "Genus", "Records", "Treatments", "Sites"))
```

```{r}
#| label: fig-yield-management
#| layout-ncol: 2
#| fig-width: 10
#| fig-height: 4.5
genus_colors <- c(Miscanthus = "red", Panicum = "blue")

nitrogen_plot <- ggplot(
  yield_nitrogen,
  aes(x = nitrogen_rate, y = yield, colour = genus)
) +
  geom_point(
    alpha = 0.35,
    size = 1,
    position = position_jitter(width = 2, height = 0, seed = 25)
  ) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  scale_colour_manual(values = genus_colors) +
  coord_cartesian(xlim = c(0, 250), ylim = c(0, 70)) +
  labs(
    x = expression("Nitrogen fertilization rate" ~ (kg ~ ha^-1 ~ yr^-1)),
    y = expression("Yield" ~ (Mg ~ ha^-1 ~ yr^-1)),
    colour = NULL,
    tag = "(a)"
  ) +
  theme(
    legend.position = "top",
    plot.tag = element_text(face = "bold")
  )

age_plot <- ggplot(
  yield_age,
  aes(x = stand_age, y = yield, colour = genus)
) +
  geom_point(
    alpha = 0.35,
    size = 1,
    position = position_jitter(width = 0.08, height = 0, seed = 25)
  ) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  scale_colour_manual(values = genus_colors) +
  coord_cartesian(xlim = c(0, 10), ylim = c(0, 70)) +
  labs(
    x = "Stand age (yr)",
    y = expression("Yield" ~ (Mg ~ ha^-1 ~ yr^-1)),
    colour = NULL,
    tag = "(d)"
  ) +
  theme(
    legend.position = "top",
    plot.tag = element_text(face = "bold")
  )

```

::: {.columns}
::: {.column width="50%"}
```{r}
#| echo: false
nitrogen_plot
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
age_plot
```
:::
:::

**Figure 4.** Yield responses to nitrogen fertilization and stand age for
*Miscanthus* and *Panicum*. Panels (a) and (d) match the corresponding panels
in the manuscript figure.

The displayed axes match the published figure for visual comparison. Axis
limits use `coord_cartesian()`, so valid observations outside the displayed
range remain in the regression summaries below.

```{r}
#| label: tbl-fig4-regressions
#| tbl-cap: "Descriptive ordinary least-squares yield regressions by genus"
summarise_lm <- function(data, predictor) {
  model <- lm(reformulate(predictor, response = "yield"), data = data)
  model_summary <- summary(model)
  slope_coefficients <- model_summary$coefficients[2, ]

  tibble(
    records = nrow(data),
    slope = unname(slope_coefficients["Estimate"]),
    std_error = unname(slope_coefficients["Std. Error"]),
    t_value = unname(slope_coefficients["t value"]),
    p_value = unname(slope_coefficients["Pr(>|t|)"]),
    r_squared = model_summary$r.squared
  )
}

regression_summary <- bind_rows(
  yield_nitrogen |>
    group_by(genus) |>
    group_modify(~ summarise_lm(.x, "nitrogen_rate")) |>
    ungroup() |>
    mutate(analysis = "Nitrogen fertilization"),
  yield_age |>
    group_by(genus) |>
    group_modify(~ summarise_lm(.x, "stand_age")) |>
    ungroup() |>
    mutate(analysis = "Stand age")
) |>
  transmute(
    analysis,
    genus,
    records,
    slope = round(slope, 3),
    std_error = round(std_error, 3),
    t_value = round(t_value, 2),
    p_value = signif(p_value, 3),
    r_squared = round(r_squared, 3)
  )

regression_summary |>
  knitr::kable(
    col.names = c("Analysis", "Genus", "Records", "Slope", "SE", "t", "p", "R-squared")
  )
```

::: {.callout-important}
## Interpreting the Figure 4 Results

These are descriptive regressions from the current public `betydata` snapshot.
They demonstrate the yield-management join and may differ from the manuscript's
2017 results because of data updates, public-data restrictions, QA/QC filtering,
and differences from the original model specification.
:::

## Geographic Distribution {#sec-geography}

```{r}
#| label: fig-geo-distribution
#| fig-cap: "Geographic distribution of Miscanthus and switchgrass yield observations"
#| fig-width: 8
#| fig-height: 5
ggplot(yield_ma, aes(x = lon, y = lat, color = genus)) +
  geom_point(alpha = 0.6, size = 2) +
  borders("world", colour = "grey70", fill = NA) +
  coord_quickmap(xlim = c(-130, 50), ylim = c(20, 70)) +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Genus"
  ) +
  theme_minimal(base_size = 12)
```

::: {.callout-tip}
## Extending This Analysis

The `sites` table contains additional geographic and climate metadata (`mat`, `map`, `soil`) that can be joined to enable climate-response analyses.
:::

## References

- LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)
